---
title: Lab Four -- Programming in R # Title of the document

# Bibliography Details
bibliography: Lab4References.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab4WriteUp.qmd")
format:
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
format-links: [pdf]
---
```{r}
#| echo: false

default_chunk_hook  <- knitr::knit_hooks$get("chunk")

latex_font_size <- c("Huge", "huge", "LARGE", "Large", 
                     "large", "normalsize", "small", 
                     "footnotesize", "scriptsize", "tiny")

knitr::knit_hooks$set(chunk = function(x, options) {
  x <- default_chunk_hook(x, options)
  if(options$size %in% latex_font_size) {
    paste0("\n \\", options$size, "\n\n", 
      x, 
      "\n\n \\normalsize"
    )
  } else {
    x
  }
})

```

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
```

* Complete the tasks below. Make sure to start your solutions in on a new line that starts with "**Solution**:".
* Make sure to use the Quarto Cheatsheet. This will make completing and writing up the lab *much* easier.

In this lab, we will build a mealkit recipe generator. I created a small website of about 40 recipes. We will scrape recipes from that website, randomly select three meals, and print a grocery list with recipe cards.

# Preliminaries 

## Part a
Below, I create a numeric vector filled with random observations. Ask for the 3rd item.<br/> 
**Note**: The `set.seed(7272)` portion ensures we all get the same answer.

```{r}
#| size: "scriptsize"
set.seed(7272) # sets the randomization seed for replication
x <- sample(x=1:10,          # sample from 1, 2, 3, ..., 10 
            size=10,         # sample of 10
            replace = TRUE)  # allowed to sample the same item multiple times
```

**Solution** 

I am going to directly grab the 3rd item using the square bracket notation. 
```{r}
#| size: "scriptsize"

(x[3])

```

## Part b

Below, I create a data frame filled with random observations.

```{r}
set.seed(7272) # sets the randomization seed for replication
df <- data.frame(x1 = sample(x=1:10,          # sample from 1, 2, 3, ..., 10 
                             size=10,         # sample of 10
                             replace = TRUE), # allowed to sample the same item multiple times
                 x2 = sample(x=1:10,          # sample from 1, 2, 3, ..., 10 
                             size=10,         # sample of 10
                             replace = TRUE), # allowed to sample the same item multiple times
                 x3 = sample(x=1:10,          # sample from 1, 2, 3, ..., 10 
                             size=10,         # sample of 10
                             replace = TRUE)  # allowed to sample the same item multiple times,
)
```

## Part b 
Use the `head(...)` function to peek at the data frame.

**Solution** 

I use head(...) here to get the first 6 rows of the data frame.
```{r}
#| size: "scriptsize"
head(df)
```

## Part c
Ask for the column `x1`.

**Solution**

I use the dollarsign notation to ask for which column I want to look at. I could also use the bracket notation if I wanted.
```{r}
#| size: "scriptsize"
(df$x1)
```

## Part d
Ask for the fifth row of the data frame.

**Solution**

I use the bracket notation to grab my specific row from the df. 
```{r}
#| size: "scriptsize"
(df[5,])
```

## Part e
Ask for the the value of `x1` in the fifth row of the data frame.

**Solution**

I use bracket notation and dollar sign notation here.  
```{r}
#| size: "scriptsize"
(df[5,"x1"])
(df[5,]$x1)
```

## Part f
Ask for the the value of in the third column and the fifth row of the data frame.

**Solution**

I use bracket notation here. I could also do dollarsign notation. 
```{r}
#| size: "scriptsize"
(df[5,"x3"])
```

## Part g
Create a sequence from 1 to 10 by 2 and use it to print the odd rows of the data frame.

**Solution**

I create my odds vector first. Then I use bracket notation wth my vector as an argument for the rows place to grab only odd rows. 
```{r}
#| size: "scriptsize"
odds = seq(1,10,2)
(df[odds,])
```

## Part h
Below, I create an empty column called `x12`. Fill in the details of the `for(...)` loop to ensure `x12` is the product of `x1` and `x2`.

**Solution**

I filled in loop making sure to have my iterator i hold place in which spot of column of x12 I am filling and using the same i to grab the right index from x1 and x2 to multiply together.
```{r}
#| size: "scriptsize"
n <- nrow(df)                # How many rows to we have to fill?
df$x12 <- rep(NA, nrow(df))   # Create an empty column for x12
for(i in 1:nrow(df)){
  df$x12[i] = df$x1[i] * df$x2[i]
}
(df$x1)
(df$x2)
(df$x12)
```

## Part i
Write a function called `calculate.score(...)` that takes three arguments representing `x1`, `x2`, and `x3`) and returns a single value based on the formula:
$$Score = (x_1 \times 2) + x_2 - x_3$$
Use your function to create a new column `total.score` in our data frame `df`.

**Solution**

Made my function taking in specifc columns. Use those columns as vectors in my function to evaluate score. put scores in its own vector. return the vector. then set the new column in the df to my returned vector. boom done. 
```{r}
#| size: "scriptsize"


calculate.score = function(x1, x2,x3){

    total.score = rep(NA, length(x1))
    for(i in 1:length(x1)){
      total.score[i] = (x1[i]*2) + x2[i] - x3[i]
    }
    return(total.score)
}
df$total.score =  calculate.score(df$x1,df$x2,df$x3)
(df$total.score)
```

## Part j
Create a function called `evaluate.row(...)` that takes one argument and returns "low" when the argument is less than 4, "mid" when the argument is between 4 and 7 (inclusive), and "high" when the arguement is 8 or larger. Then, use `sapply(...)` to apply it to column `x1`.

**Solution**

I made the function here. It uses if, else if, else to evaluate the three cases the integer could fall into. Then i use sapply to apply the function to the entire column x1 in the df. It acts like a nice for loop to keep applying the function to each number in the vector column. 
```{r}
#| size: "scriptsize"

eval.row = function(num){
  if(num<4){
    return("low")
  } else if(num>=8){
    return("high")
  }else{
    return("mid")
  }
}
(df$x1)
(sapply(df$x1,eval.row))
```

## Part k
Did we need to use loops or functions in (h.)-(j.)? That is, can we use vectorization to attain the same results in 1 line each? Where it is possible, write the line of code. Where it is not, explain why.

**Solution**

I can use the power of element-wise vector multiplication for part h. 
```{r}
#| size: "scriptsize"

(df$x12 = df$x1*df$x2)
```

Again here I can use element-wise addition, subtraction, and multiplication to recreate our previous results in part i. 
```{r}
#| size: "scriptsize"

(df$total.score = df$x1*2 + df$x2 - df$x3)

```

I can use nested ifelse conditions to alsp make my part j results into just one line of code. 
```{r}
#| size: "scriptsize"

(outcome = ifelse(df$x1>=8,"high", ifelse(df$x1<4, "low", "mid")))
```


# Complete Tasks for One Recipe

## Part a

Install and load the `rvest` package [@rvest].

**Solution**

I install and load the library here. I put eval: false, so it doesnt run every time I preview the doc. 
```{r}
#| size: "scriptsize"
#| eval: false

install.packages("rvest")
```

```{r}
#| size: "scriptsize"
library("rvest")
```

## Part b

Load the html of the `website/KimchiGrilledCheese.html` using the `read_html()` function and save the result to an object called `recipe.item`. We will use the Kimchi Grilled Cheese recipe as our prototype and extend this workflow to all recipes, so try to be as general as possible. If you do look at it, you'll notice it contains the html we saw in the developer tools in class.<br/>
**Hint**: You can use `read_html(...)` like `read.csv(...)`. Don't forget you can use the documentation to help use it.

**Solution**

Just had to make sure I was in the right working directory here. All else is good here. 
```{r}
#| size: "scriptsize"

recipe.item = read_html("website/KimchiGrilledCheese.html")
```

## Part c

Open the html file in a web browser and open developer tools. In chrome-based browsers, you can do this by pressing the three verticle dots in the upper-right corner, clicking "more tools", then "developer tools".

Find the name of the Ingredients section of the website and pull the HTML of the function using `html_element(...)` and save the results to an object called `ingredients.section`. <br/>
**Hint**: You can ask for elements by id using a preceeding "\#". See `?html_element(...)` for a helpful example. 

**Solution**

Used html_element with the id I found using developer tools.
```{r}
#| size: "scriptsize"

(ingredients.section = html_element(recipe.item, "#ingredients"))
```


## Part d
Now, we want to obtain all of the itemized items. Find the element type the individual ingredients and pull all of them from `ingredients.section` using `html_elements(...)` (note the added s) and save the results to an object called `ingredients`.<br/>
**Hint**: You can ask for elements by type by simply specifying the tag (e.g., "p" for paragraph). See `?html_element for a helpful example.

**Solution**

Found the type of the elements and searched from ingredients section. 
```{r}
#| size: "scriptsize"

ingredients = html_elements(ingredients.section,"li")
(ingredients)
```


## Part e
Similar to Part c. Find the name of the Instructions section of the website and pull the HTML of the function using `html_element(...)` and save the results to an object called `instructions.section`.<br/>

**Solution**

Same process as before here. 
```{r}
#| size: "scriptsize"

(instructions.section = html_element(recipe.item, "#instructions"))
```

## Part f
Now, we want to obtain all of the enumerated items. Find the element type the individual instructions and pull all of them from `instructions.section` using `html_elements(...)` (note the added s) and save the results to an object called `instructions`.

**Solution**

Same process as before here. 
```{r}
#| size: "scriptsize"

(instructions = html_elements(instructions.section, "li"))
```

## Part g
Find the class of the recipe image and pull the HTML using `html_element(...)` and save the results to an object called `image.element`. <br/>
**Hint**: You can ask for elements by class using a preceeding ".". See `?html_element(...)` for a helpful example. Further, note that HTML elements may have more than one class (separated by a space). When that is the case, you need to choose one.

**Solution**

Found specifc image class and used html_element. 
```{r}
#| size: "scriptsize"

(image.element = html_element(recipe.item, ".figure-img"))
```

## Part h
Use the `html_attr(...)` function to pull the source link ("src"). Then, use `paste(...)` to prepend the source link with "website/" so we have the full link.<br/>
**Note**: This would be like adding "https://www.website.com" to get the absolute link.

**Solution**

Used paste to combine "website/" and the source link. Had to make sure there is no space between the two elements in paste.  
```{r}
#| size: "scriptsize"

(image.url = paste("website/",html_attr(image.element, "src"), sep=""))
```

## Part i
Now that we have all the things we need, let's try to print the recipe. Below, I have written code to print from the objects. One by one, remove `#| eval: false` from the YAML header and add `#| echo: false` and `#| results: 'asis'`, and test. Let me know if you're stuck!

**Solution**

Followed above instructions, had no issues here. 

**Image**

```{r}
#| echo: false
#| results: "asis"
# width="50%" shrinks the image for printing
recipe.title <- recipe.item |> html_element("h1") |> html_text()
cat(paste('![', recipe.title, '](', image.url, "){width=50%}", sep="")) 
```

**Ingredients**

```{r}
#| echo: false
#| results: "asis"
formatted_list <- paste("-", html_text(ingredients))
cat(formatted_list, sep = "\n")
```

**Instructions**

```{r}
#| echo: false
#| results: "asis"
formatted_list <- paste("1.", html_text(instructions))
cat(formatted_list, sep = "\n")
```

# Complete a Full Menu!
Open `Menu.qmd` and add code to complete the following.

1. Randomly select three dinner recipes and one breakfast recipe at random.<br/>
**Hint:** Use the `sample(...)` function.
2. Pull the image, ingredients, and instructions for each selected recipe.
3. Combine all of the ingredients into a grocery list on the first page.
4. Print the image, ingredients, and instructions for each receipe on the subsequent pages.

When you render this document, no code should be visible. Instead, you should see a five-page document as described above.

# Describe your work!

## Why a fake website?
Read [https://www.scrapingbee.com/blog/is-web-scraping-legal/](Is Web Sraping Legal?). Originally, I conceived doing this with recipes from Purple Carrot (my favorite subscription service). Being a large company, their terms of service are *very* long and precluded us from copying their recipes. We also checked a few smaller recipe websites we like and even they had terms of service that restricted automated collection of data.

## Conditionals
Did you use conditional statements in your code? If yes, how. If not, are there places you could have used them but did something else?

**Solution**

I'm assuming all of these questions are in context to the code I wrote in Menu.qmd. 

I didn't use conditionals in my code. Looking back I also don't think there are any places for me to use them either as I'm not comparing any instances of data.

## Loops
Did you use loops in your code? If yes, how. If not, are there places you could have used them but did something else?

**Solution**

I did use loops in my code. I used a loop to go through each recipe and grab all the ingredients so I could combine them into one grocery list. I liked this solution because if I wanted to scale up the number of recipes, it can easily handle that. I used lapply to apply my scrape html function on each recipe in a vector, so in that case it looped through the vector. 

## Functions
Did you use functions in your code? If yes, how. If not, are there places you could have used them but did something else?

**Solution**

I tried to use functions as much as possible to cut out repeatble work. I had a scrape html function that would grab all the necessary html info given the recipe. That way, it cut out me having to repeatedly get the html four different times. I also had a print function that would nicely print out the ingredients and instructions and image in the correct format. Doing that allowed me to just call the function four times for each recipe.  


