---
title: Unit 2 Review -- tidyverse and ggplot # Title of the document

# Bibliography Details
bibliography: reviewReferences.bib  # Name of bib file in the same folder 
bibliographystyle: apa            # Bibliography style

execute:               # Default settings for code blocks
  warning: false       # Do not print warnings
  message: false       # Do not print messages
  size: "scriptsize"   # Set code size to script size

# Below will do html by default. To render the entire project
# Use the following code:
# quarto::quarto_render("Lab10WriteUp.qmd")
format:
  pdf:        # Specify .pdf formatting options 
    documentclass: article     # Document Class
    number-sections: true      # Number Sections
    reference-section-title: "References" # Title for reference section
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables 
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in
  html:       # Specify HTML formatting options 
    number-sections: true      # Number Sections
    code-fold: true            # Button to show code
    code-line-numbers: true    # Line numbers printed in code
    cap-location: top          # Captions at the top of tables and figures
format-links: [pdf]
---

## Introduction

Unit 2 focused on two tools for data analysis in R: the **tidyverse** and **ggplot2**. The tidyverse provides an easy and clean way to import, transform, and summarize data. ggplot2 provides a flexible way to build graphs that clearly communicate and summarize data. 

In this review, I will summarize the main ideas from Unit 2 and then apply them on a real dataset of professional snooker players. My goal is to create a nice guide on how and when to use the tidyverse and ggplot tools. 

## What is the tidyverse?

The tidyverse is a collection of packages made to make data science a tad easier. In class, we used tidyverse tools to read in .csv files, manipulate data, create new variables, filter observations, and compute summary statistics.

One major benefit of the tidyverse is its readability. Using the pipe operator, we can easily pass data in through steps, which makes the logic of both reading and writing the code much easier to follow. This is especially useful when cleaning data. 

## Types of variables

Another important part of Unit 2 is recognizing the different kinds of variables, because the type of variable affects both the summaries and the plots that make sense to use.

A **categorical variable** is simple and places qualitative observations into groups or categories. 

A **quantitative variable** records quantitative values. These could be either:

- **continuous**, measured data
- **discrete**, counted data

Understanding what kind of data you have is important in figuring out what kind of numerical summaries or visualizations you should provide. 

## Most important tools we used in tidyverse

- `tibble()`, to create an enhanced data frame
- `read_csv()`, to import data
- `filter()`, to keep rows based on conditions
- `mutate()`, to create or modify columns
- `group_by()`, to specify a group
- `summarize()`, to compute summaries
- `select()`, to select one or more columns as a data frame
- `case_when()`, to compute multiple if_else() statements


These tools make it easier to move from raw data to summaries in a very understandable process.

## Numerical summaries

We also learned how to describe data numerically. Depending on the type of variable and the shape of the distribution, different summaries are better.

Some of the main summaries we used were:

- mean 
- standard deviation
- median
- interquartile ranges
- counts
- proportions
- grouped summaries for comparing categories

For example, if a variable is skewed, the median and IQR may describe the center and spread better than the mean and standard deviation. If the variable is categorical, proportions are more useful than using an average.

## Introduction to ggplot2

`ggplot2` is a plotting system that drastically makes it easier to plot and graph. ggplot builds a plot by combining lines, making it easier to modularly create your graph. This usually includes:

- the dataset
- aesthetic decisions in `aes()`
- specified type of plot such as bars, points, or boxes
- axis labels
- a theme

This approach makes plots easier to customize and helps explain what each part of a graph is doing.

We also learned that plot choice depends on the type of variables involved. For example:

- one categorical variable -> bar plot
- one quantitative variable -> histogram
- one categorical and one quantitative variable -> boxplot or violin plot
- two quantitative variables -> scatterplot

This framework is useful because it gives a logical reason for choosing one graph over another, rather than just picking a graph that looks nice.

## Time to apply these skills

To review this unit, I will use a dataset on professional snooker players. This dataset includes both categorical and quantitative variables, which makes it a good fit. 

Using this dataset, I will:

- inspect the structure of the data
- create new variables with `mutate()`
- compute summary statistics
- build visualizations
- interpret what the summaries and graphs show

This will demonstrate both tidyverse data wrangling and ggplot proficiency. 

## Loading and inspecting the data

First, I want to load and read the data, so I can understand its format. 

### Load packages and data

```{r}
library(tidyverse)

data = read_csv('snooker_players_dataset.csv')
view(data)
head(data)

```

In this case, the data contains info on the player names, country of origins, rankings, match results, scoring statistics, and prize money earned. We have a nice mix of both categorical and quantitative variables, so we can practice our ggplot skills well here. 

### Select important data

```{r}
data |>
  select(Player_Name, Country, Age, World_Ranking, Matches_Won, Century_Breaks, Prize_Money_USD)

```

This is a simple example of select(). It allows me to focus on the variables I want to analyze instead of always working with the full dataset.

### Create new variables

One of the most useful tools, `mutate()`, allows me to create new columns or variables from our dataset. For example, we will make two new variables, "win_percentage" and "prize_money_millions" so that we can get eahc player's win percentage and their total prize money in the millions unit.

```{r}
data <- data |>
  mutate(
    Win_Percentage = Matches_Won / Matches_Played,
    Prize_Money_Millions = Prize_Money_USD / 1000000
  )

data |>
  select(Player_Name, Win_Percentage, Prize_Money_Millions)
```

### Numerical Summaries

Now let's use numerical summaries to better describe the distributions of our variables. 

```{r}
data |>
  summarize(
    mean_age = mean(Age),
    median_age = median(Age),
    mean_wins = mean(Matches_Won),
    median_wins = median(Matches_Won),
    mean_centuries = mean(Century_Breaks),
    median_centuries = median(Century_Breaks),
    mean_prize_money_millions = mean(Prize_Money_Millions),
    median_prize_money_millions = median(Prize_Money_Millions)
  ) |>
    pivot_longer( #fancy stuff here to display, dont worry about it
    cols = everything(),
    names_to = "Statistic",
    values_to = "Value"
  ) |>
  knitr::kable(digits = 2)
```

Now, we get a better look of the mean and medians of each variable of all the players. 

### Grouped Summary by Country

```{r}
country_summary <- data |>
  group_by(Country) |>
  summarize(
    players = n(),
    mean_age = mean(Age),
    mean_wins = mean(Matches_Won),
    mean_centuries = mean(Century_Breaks),
    mean_prize_money_millions = mean(Prize_Money_Millions)
  )
country_summary
```

This summary shows how we can use `group_by()` and `summarize()` to work together to describe stats by categories. Because this is a small data set, most countries only have 1 player, but this still represents a good example of a grouped summary. 

### Plot 1: Century breaks by player

Since `Player_Name` is categorical and `Century_Breaks` is quantitative, a bar chart is a good choice. 

```{r}
#| fig-cap: "Century breaks by player"
ggplot(data, aes(x = Player_Name, y = Century_Breaks)) +
  geom_col() +
  xlab("Player") +
  ylab("Century Breaks") +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This graph makes it easy to compare players directly. A bar chart works well here because our goal is to compare one quantitative value against multiple categories (names).

### Plot 2: Matches won and prize money

For two quantitative variables, our best option is a scatterplot. 

```{r}
#| fig-cap: Matches won and prize money earned
ggplot(data, aes(x = Matches_Won, y = Prize_Money_Millions)) +
  geom_point(size = 3) +
  xlab("Matches won")+
  ylab("Prize money (millions of USD)")+
  theme_bw()
```

This scatterplot helps show that players with more wins also tend to earn more prize money. In general, it makes sense that winning more creates more opportunities to earn more. 

### Plot 3: Boxplot of century breaks by country

Because century breaks is quntitative and country is qualitative, a boxplot is a good idea here. 

```{r}
#| fig-cap: "Century breaks by country"
ggplot(data, aes(x = Country, y = Century_Breaks)) +
  geom_boxplot() +
  xlab("Country")+
  ylab("Century Breaks")+
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This boxplot compares the distribution of century breaks across countries. Boxplots are useful because they show the center, spread, and outliers of a quantitative variable within each category. The number of players per country is small, but it still demonstrates how boxplots can be used to compare groups.

### Take aways from our analysis

We used a lot of key concepts from topic 2 here. 

First, tidyverse functions make analysis more readable and organized. I was able to move from importing the data, to selecting variables, to creating new variables, to grouped summaries in a clear process, that is easilly followed.

Second, the type of variable matters when choosing a graph. I used bar charts when comparing a quantitative measurement across players, and scatterplots when comparing two quantitative variables.

Third, `mutate()` is especially useful because it lets me create variables that are more meaningful than the originals. In this dataset, win percentage was more informative than raw wins alone, and prize money in millions was easier to interpret than large dollar amounts.

Finally, the visualizations help communicate patterns more clearly than a table alone. For example, it is easier to compare century breaks or win percentages across players in a graph than by scanning rows of numbers.

## Conclusions

Overall, this snooker dataset was a good example to test out some topics and skills from unit 2. It allowed me to combine data wrangling and data visualization in one work flow. Thank you for reading. 


